\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2016-03-07 write_up/cricket.tex
updated ms. w/ margin + split by international/domestic

The conventional wisdom among lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often the most challenging time to bat. The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours. And indeed unlimited over matches provide the greatest advantage--- the average advantage over FC and test matches is north of 2.6\% (see Figure~\ref{fig:type}). Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs. Only about 60\% of test matches end in a clear decision, the rest end in a draw. Thus, the advantage is closer to 4.5\%. The heftiest raw advantage, however, is in one-day matches (List A and ODIs), approximately 3.3\%. In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%. And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.

@@

Winning the toss ought to matter the most when the difference between the quality of the teams that are playing is the least. Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing. To study this, we collected data on team quality. The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the men's ODI teams have been published since 1981, and of the test teams, since 1952. Of the entire ranking dataset, that spans 1981--today and 1952--today for ODI and test teams respectively, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013. And scraping the latter data posed additional hurdles. We decided that the additional effort wasn't worth the small amount of additional data.}

@@

Lastly, we investigate how advantage varies by country in international matches? Are some countries better than others? We investigated the question by tallying the advantage by team. All of our estimates are imprecise enough that we cannot discount the possibility that the true effect is zero. The most puzzling result is from New Zealand, which seems to do more poorly in matches where it wins tosses than where it loses it. On the other hand, Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 3.85\% and 3.09\% respectively. Australia, Pakistan, and West Indies hover around 1--1.5\% and England is at 2.29\%.

@@

This covers some of the more straightforward questions. However, some simple questions still remain unanswered. For one, how does the toss advantage vary over time? For another, how does winning the toss affect the margin of victory?  We propose to investigate these questions in a follow-up piece.

>>>>

The conventional wisdom among lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often the most challenging time to bat. The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours. And indeed unlimited over matches provide the greatest advantage--- the average advantage over FC and test matches is north of 2.6\% (see Figure~\ref{fig:type}). Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs. Only about 60\% of test matches end in a clear decision, the rest end in a draw. Thus, the advantage is closer to 4.5\%. The heftiest raw advantage, however, is in one-day matches (List A and ODIs), approximately 3.3\%. In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%.\footnote{Splitting data by whether the match was domestic or international yields some additional insights. Like \citet{de1998winning}, who based on analysis of data from 427 international one-day matches conclude that `winning the toss at the outset of a match provides no competitive advantage' in one-day international matches, we find that in ODIs teams that win the toss win games at about the same rate as those that lose the toss. For first-class and test matches, the advantage to winning the toss is roughly the same. Meanwhile in Twenty20s, the advantage of winning the toss is greater in international than domestic matches.} And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.

@@

Winning the toss ought to matter the most when the difference between the quality of the teams that are playing is the least. Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing. To study the issue, we collected data on team quality. The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the men's ODI teams have been published since 1981, and of the test teams, since 1952. Of the entire ranking dataset, that spans 1981--today and 1952--today for ODI and test teams respectively, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013. And scraping the latter data posed additional hurdles. We decided that the additional effort wasn't worth the small amount of additional data.}

@@

Lastly, we investigate how the advantage varies by country winning the toss in international matches. Are some countries better than others at capitalizing on a toss win? We investigated the question by tallying the advantage by team that wins the toss. As ~\ref{fig:country} illustrates, all of our estimates are imprecise enough that we cannot say with confidence that any of the teams capitalizes on winning the toss. Neither can we discount the possibility that the actual differences across the teams are zero. The most puzzling result is from New Zealand. Like \citet{Saad2015}, data suggest that New Zealand does more poorly in matches where it wins tosses than where it loses it. On the other hand, Sri Lanka and India appear to do especially well, winning 3.85\% and 3.09\% additional matches, respectively, when they win the toss. Australia, Pakistan, and West Indies hover around 1--1.5\%, and England is at 2.29\%.

@@

Till now, we have focused on assessing the impact of winning the toss on the probability of winning, and how the impact is conditioned by playing conditions, by the type of match, and by the teams involved. Winning the toss, however, likely not only affects the probability of winning, but also the margin of victory. But before we assess the impact of winning the toss on the margin of victory, a short primer.

In limited over matches, when the team chasing the total falls short, the margin of victory is given in difference in runs. When the team is able to successfully chase the total, the margin of victory is the given by two numbers: number of balls remaining, and the number of wickets in hand. In unlimited overs matches, the metrics for margin of victory differ in two small ways. We don't tally the number of balls remaining when the winning team achieves the target (principally we could). Instead, we note whether or not the winning team had to bat twice---whether or not the team won by an `innings' and additional runs.

Teams that win the toss and the match in first-class and test matches win with more wickets in hand (Mean $= 6.91$) than winning teams that lose the toss (Mean $= 6.64$). In one-dayers and Twenty20s, the teams that win the toss and the match have about the same number of wickets in hand as the teams that lose the toss but win the match (One-Day: Mean$_{\text{Lose Toss}} = 5.57$, Mean$_{\text{Win Toss}} = 5.65$; Twenty20: Mean$_{\text{Lose Toss}} = 6.37$; Mean$_{\text{Win Toss}} = 6.21$).

In first-class and test matches teams that win the toss and the match also win by few more runs on average (Mean $= 136.48$; Median $=124$) than teams that lose the toss but win the match (Mean $= 133.71$; Median $=122$). Similarly, in Twenty20s, the team that wins the toss wins by a few more runs (Mean $= 37.60$; Median $=28$) than team that loses it (Mean $= 34.50$; Median $=26$). In one-dayers, however, the margin of victory is largely indistinguishable across cases where the winning team wins the toss and where it loses it (Mean$_{\text{Lose Toss}} = 63.95$, Median$_{\text{Lose Toss}} = 51$; Mean$_{\text{Win Toss}} = 63.15$, Median$_{\text{Win Toss}} = 51$).

Similar patterns hold for balls remaining---teams that win the toss generally win with a few more balls remaining than teams that lose the toss. In one-day matches, Mean$_{\text{Lose Toss}} = 49.87$, Median$_{\text{Lose Toss}} = 28$, Mean$_{\text{Win Toss}} = 53.27$, and Median$_{\text{Win Toss}} = 30$. And in Twenty20s, Mean$_{\text{Lose Toss}} = 18.69$, Median$_{\text{Lose Toss}} = 12$, Mean$_{\text{Win Toss}} = 16.41$, and Median$_{\text{Win Toss}} = 10$.

And lastly, on number of innings, the teams that win the toss win by an innings about as often as teams that lose the toss (Mean$_{\text{Lose Toss}} = 22.26\%$, Mean$_{\text{Win Toss}} = 22.43\%$).

\section*{Discussion}

The data suggest that winning the toss has a sizable impact on the probability of winning, especially in closely contested games. The data also suggest that the advantage varies considerably and systematically in expected ways, with advantage being greater in  day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match was played between closely matched teams.
